Exploitation Techniques and Mitigations

### 2016-01-22 docs/paper.tex
Paper: Draft ROP, ASLR and StackGuard

A university course at Rensselaer Polytechnic
Institut\footnote{\url{http://rpi.edu/}} held in Spring 2015 focused on
\textit{Modern Binary Exploitation}. They made their course material available
on GitHub \cite{RPISEC} under the Creative Commons Attribution-NonCommercial
4.0 International
license\footnote{\url{https://creativecommons.org/licenses/by-nc/4.0/legalcode}}.
We reused a lot of their material in this project.

@@

\section{Format String Exploits}

@@

\section{Buffer Overflow}

@@

@@

@@

\section{Stack Cookies}

>>>>

A university course at Rensselaer Polytechnic
Institut\footnote{\url{http://rpi.edu/}} held in Spring 2015 focused on
\textit{Modern Binary Exploitation}. They made their course material available
on GitHub \cite{rpisec} under the Creative Commons Attribution-NonCommercial
4.0 International
license\footnote{\url{https://creativecommons.org/licenses/by-nc/4.0/legalcode}}.
We reused a lot of their material in this project.

@@

\section{Format String Exploits}
\label{sec:format_string_exploit}

@@

\section{Buffer Overflow}
\label{sec:buffer_overflow}

@@

The first example shown in the Buffer Overflow section
(\cref{sec:buffer_overflow}) already illustrated the power that comes along
with controlling the return address. In there we were able to jump to a
completely different function upon execution. Most likely the target binary may
not contain a function doing exactly what an attacker wants to do, by
controlling the return address one can build a, so called, \textit{ROP chain}
to execute (more or less) arbitrary code, which is made out of
\textit{gadgets}.~\cite{wiki:rop,buchanan2008good,shacham2007}

\subsection{Gadgets}

A gadget is a (short) sequence of instructions ending with a \texttt{ret}.

An attacker scavenges the target binary (and used libraries) for such sequences
in order to combine them to build a malicious sequence of instructions. The
size and diversity of the code base dictates the diversity of available gadgets
and therefore the difficulty of building a specific sequence.

The shell code used in this paper could be mapped to three gadgets, for
example, if the described gadgets were available. The first one pushes the
string \texttt{"/bin//sh"} onto the stack, the second one sets up the registers
(arguments) and the last one calls the \texttt{execve} system call.

The final \texttt{ret} statement of a gadgets is required for chaining them
together. We not only use the Buffer Overflow to control one, but multiple
return addresses. Because of this our payload will be made up by a list of
return addresses (the start of each gadgets) interleaved by some padding. Each
provided address will be consumed one by one at the end of each gadgets to
\emph{jump} to the next one.

\begin{listing}[h!]
    \codefile{../src/return_oriented_programming/gadgets.txt}
    \caption{Finding available gadgets in a binary}
    \label{src:rop_gadgets}
\end{listing}

The easiest way to find what gadgets are available may be by piping the output
of \texttt{objdump} to \texttt{grep} filtering for \texttt{ret} statements.
This is done in \cref{src:rop_gadgets}. In there you can see three possible
gadgets with different instructions. Gadgets can of course be arbitrary long,
we just used a length of 5 instructions for this example.

There are algorithms and tools available which simplify the process of finding
such gadgets (and even whole chains) but they go beyond the scope of this
writing.~\cite{shacham2007}

\subsection{Example}

This example is taken from a blog
post\footnote{\url{http://codearcana.com/posts/2013/05/28/introduction-to-return-oriented-programming-rop.html}}
on Code Arcana, which also includes a simpler as well as a more complex example
about ROP.

\begin{listing}[h!]
    \cfile{../src/return_oriented_programming/rop.c}
    \caption{Example for exploiting a Buffer Overflow with ROP}
    \label{src:rop_example}
\end{listing}

The target program is displayed in \cref{src:rop_example}. We will not be able
to inject and execute shell code, and there is no function present which
directly opens up a shell for us. But there are parts which can be used to do
so.

\newpage

% TODO

@@

This mitigation technique was introduced to render ROP (and ret2libc) void. The
idea behind it is quite simple, and the name gives it away already. Memory
layout is randomize so an attacker cannot reliably use ROP. An attacker will
not be able to copy the exact setup of a target machine by only knowing which
binary (and libraries) is used.

\begin{listing}[h!]
    \begin{code}
        > echo 2 > /proc/sys/kernel/randomize_va_space

        > cat /proc/self/maps
        08048000-08054000 r-xp 00000000 08:01 131085     /bin/cat
        08054000-08055000 r--p 0000b000 08:01 131085     /bin/cat
        08055000-08056000 rw-p 0000c000 08:01 131085     /bin/cat
        091de000-091ff000 rw-p 00000000 00:00 0          [heap]
        b7531000-b76e5000 r-xp 00000000 08:01 917531     /lib/i386-linux-gnu/libc-2.21.so
        b76f7000-b7719000 r-xp 00000000 08:01 917507     /lib/i386-linux-gnu/ld-2.21.so
        bfe0d000-bfe2e000 rw-p 00000000 00:00 0          [stack]

        > cat /proc/self/maps
        08048000-08054000 r-xp 00000000 08:01 131085     /bin/cat
        08054000-08055000 r--p 0000b000 08:01 131085     /bin/cat
        08055000-08056000 rw-p 0000c000 08:01 131085     /bin/cat
        093e3000-09404000 rw-p 00000000 00:00 0          [heap]
        b7560000-b7714000 r-xp 00000000 08:01 917531     /lib/i386-linux-gnu/libc-2.21.so
        b7726000-b7748000 r-xp 00000000 08:01 917507     /lib/i386-linux-gnu/ld-2.21.so
        bf962000-bf983000 rw-p 00000000 00:00 0          [stack]

        > cat /proc/self/maps
        08048000-08054000 r-xp 00000000 08:01 131085     /bin/cat
        08054000-08055000 r--p 0000b000 08:01 131085     /bin/cat
        08055000-08056000 rw-p 0000c000 08:01 131085     /bin/cat
        094ec000-0950d000 rw-p 00000000 00:00 0          [heap]
        b7588000-b773c000 r-xp 00000000 08:01 917531     /lib/i386-linux-gnu/libc-2.21.so
        b774e000-b7770000 r-xp 00000000 08:01 917507     /lib/i386-linux-gnu/ld-2.21.so
        bfb24000-bfb45000 rw-p 00000000 00:00 0          [stack]
    \end{code}
    \caption{Let \texttt{cat} show its memory mappings with ASLR enabled (some
lines have been omitted)}
    \label{src:aslr_cat}
\end{listing}

ASLR is enabled by default and one can easily check the implications by running
\texttt{cat} on \texttt{/proc/self/maps} a few times as shown in
\cref{src:aslr_cat}. Line 10, 19 and 28 show, for example, that the stack
starts at different locations in memory each time cat has been invoked.

We can directly see one flaw in this setup --- not all sections of the
\texttt{cat} binary start at random locations. Especially the \texttt{.text}
always starts at the same position. This happens because \texttt{cat} itself
was not compiled as Position Independent Executable (PIE). Since this is
actually the default behaviour, most programs' \texttt{.text} segment will
always start at the same location. One could pass the corresponding flag
(\texttt{-pie}) to the compiler to prevent this, then ASLR would be able to
randomize these segments too.

Breaking ASLR, even when the code is compiled with \texttt{-pie}, is easier
than it seems at first. Relocation only happens to a section at whole, inside a
section still shares the same relative distance as it would without ASLR. But
before exploit this fact have a look at the randomize addresses again.

Only three nybble ($3 \times \SI{4}{\bit}$) differ between multiple runs giving
us $2^{12} = 4096$ possibilities. If the scenario allows it, brute forcing is a
viable option here. But note that this changes drastically for \SI{64}{\bit}.

@@

Nevertheless a better option has already been teased. ASLR can be broken easily
as soon as we get \emph{one} address (pointer) to a section of interest is
\emph{leaked}. Therefore the name information leak. We show the implications of
such a leak by an example taken from \cite{rpisec}.

\begin{quote}
    Lets say you managed to leak a pointer (\texttt{0xb7e72280}) and you know
    that this one usually points to \texttt{printf}.
    \medskip

    Look how far away \texttt{system} is from \texttt{printf}, in the standard
    library. It's \texttt{0xd0f0} bytes.
    \medskip

    We now know that \texttt{system} is at:
    \[ \mathtt{0xb7e72280} - \mathtt{0xd0f0} = \mathtt{0xb7e65190} \]
\end{quote}

In case you may wonder how easy it is to leak information like a pointer, this
already happened to us as side effect back at the format string example
(\cref{sec:format_string_exploit}).

The adaptation for the exploit we were using looks as follows. First, manage to
leak a pointer somehow which enables you to calculate the address offset
introduced by ASLR. Augment your ROP chain to take the offset into account. Run
the exploit. Since this rather simple and we already gave an example how to
calculate the offset, we would like to leave this as an exercise for the
reader.

Adapt the target file used in the ROP example by printing the address of
\texttt{printf} first and then read in the payload via \texttt{stdin}. This way
you can first simulate a leaked pointer, adapt the ROP chain and run it. Double
check the distances between functions in the libraries, they may not correlate
with the ones used in this writeup.

\section{StackGuard}

DEP can be fooled by ROP and ASLR is rendered useless with a simple info leak.
Something else is required at this point. Thinking back, the original
vulnerability was the overwrite of the return address located on the stack. Two
(additional) counter mechanisms were introduced going by the names StackGuard
and StackShield. We will take only a look at StackGuard and one relative common
scenario to break it, but there is a comprehensive article
\cite{phrack_stack_guard} on \texttt{phrack}\footnote{\url{http://phrack.org/}}
describing and breaking both techniques.

The general idea behind StackGuard is to place \emph{something} before the
return address which \emph{guards} against buffer overflows. This
\emph{something} known as a Canary and comes in different forms.

\begin{description}
    \item[Terminator] A terminator Canary contains a sequence of commonly used
        terminator symbols (like null, EOF, linefeed, \dots) to
        \emph{terminate} most string operations. Hence the return address will
        not be overwritten.
    \item[Random] A random Canary is chosen at program start and stored
        \textit{somewhere save} and pushed into the stack upon function calls
        to \emph{guard} the return address. The canary on the stack is checked
        against the one stored save before issuing the return. The program is
        terminated upon mismatch. With this setup an attacker has to know the
        canary in order to overwrite the return address. Since it is chosen at
        random when the program starts, he cannot reliably reproduce the same
        canary in his cloned setup.

        In our case the original canary will be stored in one of the segment
        registers\footnote{\url{https://en.wikipedia.org/w/index.php?title=X86_memory_segmentation&oldid=697253060}
        see \textit{Later developments}}.
\end{description}

There is also the random XOR Canary which XORs the random Canary with the
return address before placing it on the stack. ``This is effectively encryption
of the return address with the random canary of this
function.''~\cite{phrack_stack_guard}

The practical approach is taken next by looking at the stack frame of a
vulnerable function when compiled without \texttt{-fno\_stack\_protector}.

\begin{listing}[h!]
    \begin{minipage}[t]{0.32\textwidth}
        \cfile{../src/canary/vuln.c}
    \end{minipage}
    \begin{minipage}[t]{0.33\textwidth}
        \codefile{../src/canary/vuln_gdb1.txt}
    \end{minipage}
    \begin{minipage}[t]{0.33\textwidth}
        \codefile{../src/canary/vuln_gdb2.txt}
    \end{minipage}
    \caption{Examining the canary as generated by GCC}
    \label{src:canary_gcc}
\end{listing}

As can be seen in \cref{src:canary_gcc} the buffer was not filled beyond his
capacity to examine the Canary located in the same stack frame. A script
created by Daniel Walter\footnote{\url{http://0x90.at/post/gdb-stack-script}}
has been adapted slightly to display the stack together with some annotations.
Using \texttt{"AAAAAAA\textbackslash{}n"} and
\texttt{"BBBBBBB\textbackslash{}n"} makes the buffer clearly visible in lines
16 and 17. The Canary can be observed in line 18.

The Canary itself is composed of a terminator (null) followed by a random
sequence of \SI{3}{\byte}. This sequence changes very time the program is run.
Feeding more data to the buffer and overflowing it this way yields termination
of the program. Note that the \texttt{puts} function is still called, the
termination happens just before the return of the function.

\begin{code*}{linenos=false}
    > ./vuln <<< "AAAAAAAADEADBEEF"
    AAAAAAAADEADBEEF

    *** stack smashing detected ***: ./vuln terminated
    Aborted (core dumped)
\end{code*}
